Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-29 text/supp.tex
Edit edit edit

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
expect values of $b$ with levels of density dependence as low as $-1$ (very
strong inverse density dependence), nor would we generally expect values above $1$. We
allowed values of $b$ above $1$ to allow for somewhat non-stationary time
series of growth rates. The estimates of $b$ were well within these bounds. Our
prior on $\lambda$ was very weakly informative within the range of expected
values for population growth and is similar to the default priors suggested by
\citet{gelman2008d} for intercepts of regression models. Our prior on $\sigma$
follows \citet{gelman2006c} and \citet{gelman2008d} and is based on our
expected range of the value in nature from previous studies
\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to an
uninformative uniform prior on $\sigma$, but the weakly informative Half-Cauchy prior
substantially sped up chain convergence.

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose a more
informative exponential rate parameter of $0.1$. We chose a less informative
rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
< 2$ the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the
scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
the posterior will approximately match the prior (prior median $= 71$, mean $= 102$)
and not be flagged as likely heavy tailed using the metrics we used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$).

We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the scale
reduction factor --- a measure of chain convergence) was greater than $1.05$
for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a
measure of the effective number of uncorrelated samples) for any parameter was
less than $200$, we doubled both the total iterations and warm up period and
sampled from the model again. These thresholds are in excess of the minimums
recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size
$> 100$ for reliable point estimates and confidence intervals. In the majority
of cases our minimum thresholds were greatly exceeded. We continued this
procedure up to $8000$ iterations ($16000$ total samples) by which all chains
were deemed to have sufficiently converged. These chain lengths may seem low to
those familiar with software such as WinBUGS or JAGS, but the No-U-Turn
Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far
fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.

We separated our simulation into these two components to avoid confounding two
issues. (1) With smaller sample sizes, there may not be a stochastic draw from
the tails of a distribution. In that case, no model, no matter how perfect,
will be able to detect the shape of the tails. (2) Complex models may return
biased parameter estimates if there are conceptual, computational, or coding
errors. Our first simulation tested the first issue and our second simulation
tested the latter. In general, out simulations show that, if anything, our
model under predicts the magnitude and frequency of heavy tailed events ---
especially given the length of the time series in the GPDD.

\subsection{Estimating $\nu$ simulations}

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
usually qualitatively distinguish heavy from not-heavy tails, but the
model tended to underestimate how heavy the tails were. At the same time, the
model tended to overestimate how large the scale parameter was
(Fig.~\ref{fig:sim-gompertz} lower panels).

Our results show that the more-complicated Gompertz model can still recapture
the true value of $\nu$ when the process noise was chosen so that appropriate
tail events were present (Figs~\ref{fig:sim-gompertz} and
\ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
Likewise, the other Gompertz parameters were estimated without any systematic
bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
green symbols). The addition of observation error caused the model to tend to
underestimate the degree of heavy-tailedness, overestimate the magnitude of
process noise, somewhat overestimate $\lambda$, and overestimate density
dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with
observation error is a known phenomenon \citep{knape2012}. Fitting a model with
correctly specified observation error made marginal improvements to model
bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
\ref{fig:sim-gompertz-boxplots}).

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
models distinguished heavy and not-heavy tails reasonably well. Without
observation error, and using a probability of $0.5$ as a threshold, the model
would have miscategorized only one of $40$ simulations at $\nu = 5$ across
a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper
panels). The model would have correctly categorized all cases where the process
noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in
Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$. With $0.2$ standard
deviations of observation error, the model still categorized XX\% of cases as
heavy tailed when $\nu = 5$ and all cases where $\nu = 3$. Allowing for
observation error made little improvement to the detection of heavy tails
(Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore, we
chose to focus on the simpler model without observation error in the main
text, particularly given that the true magnitude of observation error was
unknown in the empirical data.

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fitted to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an
additional parameter $\phi$, which represents the relationship between
subsequent residuals. Based on the results of previous analyses with the GPDD
\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
running our model across all populations, we placed a weakly informative prior
on $\phi$ that assumed the greatest probability density near zero with the
reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
\mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did
not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
> 200$) after 8000 iterations of four chains. We did not include these models in
Fig.~\ref{fig:alt}.

@@

@@

We fitted a multilevel beta regression model to the predicted probability of
heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
dynamics. The beta distribution is useful when response data range on
a continuous scale between zero and one. We used a logit link function as is
typically used in logistic regression. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,\\
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
represents the predicted value for population $i$, class $j$, order $k$, and
species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
represents a vector of predictors (such as lifespan) for population $i$ with
associated $\beta$ coefficients. The intercepts are allowed to vary from the
overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
taxonomic order ($\alpha^\mathrm{order}_k$), and species
($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
  \mathrm{species}}$. Where possible, we also allowed for error distributions
around the predictors by incorporating the standard deviation of the posterior
samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the
mean point value as normal distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on
all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the
dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all
other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
the Cauchy priors concentrate more probability density at reasonably parameter
values while allowing for higher probability density far into the tails, thereby
allowing the data to dominate the posterior more strongly if it disagrees with
the prior. Our conclusions were not qualitatively changed by using uniform
priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up
iterations, four chains, and discarding every second sample to save memory. We
checked for chain convergence visually and with the same criteria as before
($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails
accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted
a separate multilevel model with the same structure but with only time-series
length as a predictor. (In this case, we did not want to control for intrinsic
population characteristics such as density dependence.) Since our predictors were
centered by subtracting their mean value, we obtained order-level
estimates of the probability of heavy tails at mean time-series length by
adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
$\alpha^\mathrm{order}_k$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/priors-gomp-base.pdf}
\caption{
  Probability density of the Bayesian priors for the Gompertz models. From left
  to right: (1) per capita growth rate at $\ln$(abundance) = $0$: $\lambda \sim
  \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
  noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0,
  2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1
  correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen
    Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$,
  the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
}
  \label{fig:priors}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
  the process deviations (bottom panels) for a given number of samples (columns)
  drawn from a distribution with a given true $\nu$ value (rows). The red lines
  indicate the true population value. When a small number of samples are drawn
  there may not be samples sufficiently far into the tails to recapture the
  true $\nu$ value; however, heavy tails are still distinguished from normal
  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
  = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
\caption{
  Simulation testing the Gompertz estimation model when the process deviation
  draws were chosen so that $\nu$ can be estimated close to the true value
  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
  specified $\nu$). The simulation was run across population $\nu$ values
  (columns) and different scenarios (colours): (1) 100 time steps and no
  observation error, (2) 50 time steps and no observation error, (3) 50 time
  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
  ignored, and (4) 50 time steps with observation error in which the quantity
  of observation error was assumed known. Dashed horizontal lines show the true
  population values; these true values were chosen to represent approximately
  the median values as estimated from the GPDD. Individual dots and lines
  represent a stochastic draws from the true population distribution and
  a model fitting (median and 80\% credible intervals). The panels from top to
  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
  parameter $\lambda$, and the density dependence parameter $b$.
}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
\caption{
  The distribution of median estimates from the same simulation as described in
  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
  whiskers extend to $1.5$ times the interquartile range and outliers are shown
  as dots.
}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
\caption{
  Time series for populations with p($\nu < 10$) $>$ 0.5 using the base
  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO
  add dots for upswings and downswings as in Table 1. The labels on each panel
  indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD
  ID number.
}
\label{fig:heavy-ts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
\caption{
  Estimates of $\nu$ from alternative models plotted against the base Gompertz
  model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\%
  credible intervals (segments). The diagonal line indicates a one-to-one
  relationship. Different colours indicate various taxonomic classes. The
  grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken
  as a threshold of heavy-tailed dynamics. The Gompertz observation error model
  assumes a fixed standard deviation of observation error of $0.2$ on the log
  scale.
}
\label{fig:alt}
\end{center}
\end{figure}

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error:

@@

\noindent
The GPDD IDs used in our analysis:

>>>>

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
expect values of $b$ with levels of density dependence as low as $-1$ (very
strong inverse density dependence), nor would we generally expect values above
$1$. We allowed values of $b$ above $1$ to allow for non-stationary time series
of growth rates. The estimates of $b$ were well within these bounds. Our prior
on $\lambda$ was very weakly informative within the range of expected values
for population growth and is similar to the default priors suggested by
\citet{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior
on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and the scale
parameter of $2.5$ is based on our expected range of the value in nature from
previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample of
populations, our parameter estimates were not qualitatively changed by
switching to an uninformative uniform prior on $\sigma$, but the weakly
informative Half-Cauchy prior substantially sped up chain convergence.

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
exponential rate parameter of $0.1$. We chose a less informative rate
parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
to avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the scenario
where the data are informative about heavy tails
(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
tailed.

We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
scale reduction factor --- a measure of chain convergence) was greater than
$1.05$ for any parameter or the minimum effective sample size,
$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
for any parameter was less than $200$, we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
and effective sample size $> 100$ for reliable point estimates and confidence
intervals. In the majority of cases our minimum thresholds were greatly
exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
samples) by which all chains were deemed to have sufficiently converged. These
chain lengths may seem low to those familiar with software such as WinBUGS or
JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
generally requires far fewer iterations to obtain equivalent effective sample
sizes \citep{stan-manual2014}.

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fitted to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an
additional parameter $\phi$, which represents the dependence of
subsequent residuals. Based on the results of previous analyses with the GPDD
\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
running our model across all populations, we placed a weakly informative prior
on $\phi$ that assumed the greatest probability density near zero with the
reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
\mathrm{max.} = 1)$. The MCMC chains did
not converge for \modelsNoConvergeAROne\ populations according to
our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
> 200$) after 8000 iterations of four chains. This included only
\modelsNoConvergeAROneHeavyBase\ populations in which
Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
We did not include these models in Fig.~\ref{fig:alt}.

@@

We separated our simulation into these two components to avoid confounding two
issues. (1) With smaller sample sizes, there may not be a stochastic draw from
the tails of a distribution. In that case, no model, no matter how perfect,
will be able to detect the shape of the tails. (2) Complex models may return
biased parameter estimates if there are conceptual, computational, or coding
errors. Our first simulation tested the first issue and our second simulation
tested the latter. In general, our simulations show that, if anything, our
model under predicts the magnitude and probability of heavy tailed events ---
especially given the length of the time series in the GPDD.

\subsection{Estimating $\nu$ from a stationary t distribution}

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~\ref{fig:sim-gompertz} lower panels).

Our results show that the Gompertz model can recapture the true value of $\nu$
when the process noise was chosen so that appropriate tail events were present
(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green
symbols in the top rows). Likewise, the other Gompertz parameters were
estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
\ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of
observation error caused the model to tend to underestimate the degree of
heavy-tailedness, overestimate the magnitude of process noise, somewhat
overestimate $\lambda$, and overestimate density dependence (blue symbols in
Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The
overestimation of density dependence with observation error is a known
phenomenon \citep{knape2012}. Fitting a model with correctly specified
observation error made marginal improvements to model bias (purple symbols in
Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
models distinguished heavy and not-heavy tails reasonably well. Without
observation error, and using a probability of $0.5$ as a threshold, the model
correctly classified all simulated systems with normally distributed process
noise as not heavy tailed. The model would have miscategorized only one of $40$
simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points
(Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly
categorized all cases where the process noise was not heavy tailed (indicated
as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu
= 3$. With $0.2$ standard deviations of observation error, the model still
categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and
all cases where $\nu = 3$. Allowing for observation error made little
improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob},
lower-left vs.\ lower-right panel). Therefore, we chose to focus on the simpler
model without observation error in the main text, particularly given that the
true magnitude of observation error was unknown in the empirical data.

@@

We fitted a multilevel beta regression model to the predicted probability of
heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
dynamics. The beta distribution is useful when response data range on
a continuous scale between zero and one. We used a logit link function as is
typically used in logistic regression. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,\\
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
represents the predicted value for population $i$, class $j$, order $k$, and
species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
represents a vector of predictors (such as lifespan) for population $i$ with
associated $\beta$ coefficients. The intercepts are allowed to vary from the
overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
taxonomic order ($\alpha^\mathrm{order}_k$), and species
($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
  \mathrm{species}}$. Where possible, we also allowed for error distributions
around the predictors by incorporating the standard deviation of the posterior
samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the
mean point value as normal distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on
all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the
dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all
other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
the Cauchy priors concentrate more probability density around expected parameter
values while allowing for a higher probability density far into the tails, thereby
allowing the data to dominate the posterior more strongly if it disagrees with
the prior. Our conclusions were not qualitatively changed by using uniform
priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up
iterations, four chains, and discarding every second sample to save memory. We
checked for chain convergence visually and with the same criteria as before
($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails
accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted
a separate multilevel model with the same structure but with only $\log$
time-series length as a predictor. (In this case, we did not want to control
for intrinsic population characteristics such as density dependence.) Since our
predictors were centered by subtracting their mean value, we obtained
order-level estimates of the probability of heavy tails at mean log time-series
length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
$\alpha^\mathrm{order}_k$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/priors-gomp-base.pdf}
\caption{
  Probability density of the Bayesian priors for the Gompertz models. From left
  to right: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim
  \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
  noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0,
  2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1
  correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen
    Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$,
  the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
}
  \label{fig:priors}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
  the process deviations (bottom panels) for a given number of samples (columns)
  drawn from a distribution with a given true $\nu$ value (rows). The red lines
  indicate the true population value. When a small number of samples are drawn
  there may not be samples sufficiently far into the tails to recapture the
  true $\nu$ value; however, heavy tails are still distinguished from normal
  tails in most cases, even with only 25 or 50 samples.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
\caption{
  Simulation testing the Gompertz estimation model when the process deviation
  draws were chosen so that $\nu$ could be estimated close to the true value
  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
  specified $\nu$). The simulation was run across population $\nu$ values
  (columns) and different scenarios (colours): (1) 100 time steps and no
  observation error, (2) 50 time steps and no observation error, (3) 50 time
  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
  ignored, and (4) 50 time steps with observation error in which the quantity
  of observation error was assumed known. Dashed horizontal lines show the true
  population values; these true values were chosen to represent approximately
  the median values as estimated from the Global Population Dynamics Database.
  Individual dots and lines
  represent a stochastic draws from the true population distribution and
  a model fitting (median and 80\% credible intervals). The panels from top to
  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
  parameter $\lambda$, and the density dependence parameter $b$.
}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
\caption{
  The distribution of median estimates from the same simulation as described in
  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
  whiskers extend to $1.5$ times the interquartile range and outliers are shown
  as dots. Note that these distributions represent the output from only 20
  simulation iterations, so there is some level of remaining sampling noise.
}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
\caption{
  Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
  labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the
  species, and the Global Population Dynamics Database ID number.
}
\label{fig:heavy-ts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
\caption{
  Estimates of $\nu$ from alternative models plotted against the base Gompertz
  model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\%
  credible intervals (segments). The diagonal line indicates a one-to-one
  relationship. Different colours indicate various taxonomic classes. The
  grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken
  as a threshold of heavy-tailed dynamics. The Gompertz observation error model
  assumes a fixed standard deviation of observation error of $0.2$ on a log
  scale.
}
\label{fig:alt}
\end{center}
\end{figure}

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error. The specific code for used for the various models in our analysis is available at \url{https://github.com/seananderson/heavy-tails}.

@@

\noindent
The GPDD IDs used in our analysis. See \url{https://github.com/seananderson/heavy-tails}.
